### 2009-12-04 docs/marcello-dissertation/thesis/chapter2-state-of-the-art.tex
Print reviews of chapter 1 and 2, as suggested by Kleber. It definitely helped to see things we can't see when writing on the computer screen. :)
git-svn-id: svn://localhost/netbeams/branches/marcello/persistence@668 559edde2-a4d4-44e0-96fc-4035e40d7396

In general, this chapter reviews sensor networks characteristics and
properties, necessary to the understanding of the basic properties of a sensor
network. Moreover, it reviews the characteristics of data persistence for
sensor networks, characterizing the different approaches used to store data
such as data model used and how the collected data is described.

\section{Survey in Sensor Networks and Its Data}

Sensor Networks are specialized types of network systems comprised of nodes
with specific goals such as to produce and collect data. The former is seen
as a sensor device capable of producing data to the network based on an
observation from the environment or just a raw value regarding a given
propertiy, while the latter can be seen as the network sink responsible to
receive the observed or ``sensed'' data. In this way, data persistence in sensor
networks depends on the approaches used to store the data on the data sink,
since these types of nodes can be located in the network close to the devices 
or in a centralized location. As a consequence, the query used to query the
data can take place in the network nodes themselves or in the centralized node.
The former approach is used when the data is collected at the sensor node on
its local storage \cite{sn-storage01, sn-storage03}, strategy typically
used to maximize the sensor's resources, while minimizing energy consumption
\cite{sn-storage04}.The latter strategy, however, has been used when the
collected data is archived in a centralized database system for later reuse
\cite{sn-storage02}.

When it comes to data representation, the relational data model
\cite{relational-model} is one of the most used approaches to represent data in
sensor network databases in any of the query processing strategies and, in most
cases, with a modified version of the SQL language \cite{sn-db-newop}. However,
the design of the database system must be done with prior knowledge of the
sensor types when choosing the relational data model, as the database design
must be normalized \cite{db-normalization} for the entities already identified
for the system. For example, the introduction of a new sensor device might
represent a potential change on the database structure by seeking a new
normalized version of the database design. In addition to challenges of
maintaining the database schema normalized, the use of Data Provenance 
approaches in sensor network data usually addresses the problem related with
the origin of data and its entire history \cite{sn-provenance}.

The following subsections describe the main characteristics of sensor networks
described in the literature, and then others that interest data persistence in
sensor networks.

\subsection{Deployment and Mobility}

Each node of the sensor network can be deployed in an arbitrary method such as
being dropped from an airplane, or in a manually placed at chosen geographic
locations \cite{sn-intro01}. Moreover, the deployment can be done as a one-time
event or as an iterative process, depending on the needs of where the sensor
device needs to be placed.

Nodes can be able to move around the deployed locality or remain fixed at a
single place. In the case of a sensor networks comprised with different moving
sensors, a given sensor node can move occasionally or continuously, as a result
of an accidental side-effect or a desired property of the device. Usually,
these sensors are automotive or they can be attached to another moving object.
For this reason, \cite{sn-intro01} describes those types of mobility as active
or passive.

\subsection{Device Size, Resources, Cost, Energy, Heterogeneity}

Such sensor nodes can be of different sizes, ranging from microscopically small
particles, small devices to bigger ones with a few feet cubic. In addition to
being limited in size, they are also limited in resources and computational
power, as they are designed to either stored energy on batteries or
use energy scavenged from solar cells \cite{sn-intro01}.

Sensor devices range in size, capabilities and other properties, which are
associated to their cost ranging from hundreds to thousands of dollars. All in
all, the price depends on the device's functionality, energy utilization
capabilities, among others. For this reason, \cite{sn-intro01} explains that a
sensor network may potentially contain similar sensor devices, as they can
contain dissimilar ones from each other, which characterizes the heterogeneity
of a sensor network by the hardware capabilities such as GPS enabled or having
GSM network connectivity.

\subsection{Communication Mode, Network Infrastructure, Coverage and
Connectivity}
\label{sec:sn-infrastructure}

Sensor nodes can communicate in different modes such as radio, diffuse
light, inductive or sound, which may influence on the type of protocols used to
communicate among the nodes, where they can be said to be in an organized
infrastructure or as an ad hoc way \cite{sn-intro01}.

When it comes to infrastructure, sensor networks differ in topology:
single-hop, star, networked stars, tree and graph are examples of these
topologies. Each of they topologies may influence on how data is collection
and routed to their data sink. While there are only two levels of node
connectivity in sensor networks deployed as single-hop and single start
topologies, networks whose topology is either tree, graph or interconnected
stars use communications among each of their nodes for data communication.

Last, but not least, the coverage of a sensor network is related to the
physical space that is ``visible" to the sensors, which is directly related to
the purpose of the network, its size in terms of the number of nodes that are
deployed and the sensor network lifetime. In this way, sensor devices can be
deployed in different areas and distributed in sparse or dense manners, where
the sensor network they support is providing the necessary data that
represents the coverage of the network. However, considering that some types of
sensor devices are very likely to regular hardware failures, projects have
provided redundant nodes \cite{sn-intro01} to provide better data coverage.

No matter the type of coverage provided by a sensor network, different levels
of coverage may also influence on the connectivity of the sensor nodes. In
order to provide connectivity to redundant nodes, for example, a physical 
connection must be available at all times. On the other hand, sensor nodes can
also communicate with other nodes in a sporadic way when they are designed to
be isolated with only communication activities to report the collected data to
their immediate peer (other nodes and/or data sink).

\subsection{Network Size, Lifetime and Quality-Of-Service}

The network size is related to the degree of coverage of a sensor network, and
it is measured by the number of devices deployed as nodes of the network.
\cite{sn-intro01} reports examples of sensor networks with sensor nodes ranging
from dozens, hundreds or even thousand of network nodes.

The application of the sensor network determines for how long the sensor network
may be exist. For example, the existence of an environmental sensor networks
\cite{sn-ex01} may be not determined since its the data collected from the
network are used for archival purposes, hence, without termination of the
execution of its sensing tasks. On the other hand, however, the lifetime of a
sensor network designed to count the number of people who entered in different
rooms of a building during an event has its life-time related to the period of
the event.

Finally, the design of sensor networks can be related to the constraints
related to quality-of-service (QoS) such as real-time delivery within a
certain period of time or robustness such as maintaining data alive even
though there a communication link is not available \cite{sn-intro02}.

@@

The most valuable objective of a sensor network is related to its first
product: the use of the raw data. Data collection in sensor networks is
directly dependent on the specifications of the sensor network, its routing
mechanisms and the final destination of the produced data. In a nutshell,
\cite{sn-storage03} describes data as follows:

\begin{itemize}
  \item \textbf{low-level property} as the raw data collected from the sensor
  device; For example, the sea level as a number;
  \item \textbf{Observation or Event} as the processed data in the format of
  the final observation. For instance, if the ocean is in its highlight or low
  tight.
\end{itemize}

\subsection{How the Collected Data is Used}
\label{sec:sn-data-purpose}

At first glance, the raw data serves as the main outcome of the sensor network.
However, the first consideration that must be considered is regarding the
purpose of the collected data, as \cite{sn-provenance} explains differentiates
between two main benefit of the sensor data:

\begin{itemize}
  \item \textbf{Real-time Data Stream}, as the generated data from the sensor
  is not transmitted, but extracted from the device as it is requested. The
  data is still in its standard form specified by the sensor manufacturer;
  \item \textbf{Archival Data}, as the collected data include historical
  significance or will be use by other applications at a later time.
\end{itemize}

While real-time data stream are used and can potentially be discarded, the
archival data needs to be stored in a secondary memory. Nonetheless, given the
nature of the sensor device specification, the low-level data produced by a
given sensor device might not include vital information regarding the identity
of the data itself. The use of Data Provenance techniques is described by
\cite{sn-provenance} as an essential step to provide the collected data
specific metadata regarding its location and lineage.

Under these circumstances, the collected data needs first to face its journey
starting from its transmission from its producing device to the specified data
sink as specified by \cite{sn-storage01, sn-storage02, sn-storage03}.

\subsection{Storage Location for Collected Data}
\label{sec:sn-storage-locations}

Given that collected data can be used for different purposes as described in
section \ref{sec:sn-data-purpose}, the sensor node that produces the data is
responsible to send the raw data to a specified data storage target. Upon
producing data, a sensor network node is specified to deal with the produced
data where, in most cases, depending on the sensor network infrastructure used
\cite{sn-storage03} (see section \ref{sec:sn-infrastructure}). In such a way,
different storage mechanisms are provided:

\begin{itemize}
  \item \textbf{External Storage}: the sensor device sends the produced data to
  another sensor node that contains an external storage.
  \item \textbf{Local Storage}: as it is the case of real-time data stream,
  where data is allocated on the sensor device's main read-only memory.
\end{itemize}

It is clear that the data flows from the sensor node to another depending on
the definition of the network. On the whole, the sensor network infrastructure
dictates where the data will flow to, as described in section
\ref{sec:sn-infrastructure}. \cite{sn-storage02} sees a wireless sensor
networks as a many-to-one data collection pattern, when its organization is
based on single-hop or star. On the other hand, when the data organization is
structured as a graph, tree, etc, the data flows from one intermediate sensor
node to another until it reaches its final storage node as described by
\cite{sn-storage01, sn-storage03}.

Last, but not least, yet a different storage strategy can be used: Data-Centric
Storage. This strategy aims at clustering the collected data based on a given
nature of the data such as location. Regardless the data storage mechanism
used, the query processing will depend on the sensor network infrastructure and
data location.

\subsection{Query Processing for Storage Locations}
\label{sec:query-process}

Given the different locations where the collected data might be as described in
section \ref{sec:sn-storage-locations}, different query processing for the
collected sensor data can be used as described by \cite{sn-storage03}:

\begin{itemize}
  \item \textbf{In-Network Query Processing}: sensor nodes that provide
  real-time data stream usually replies to requests to properties.
  Likewise, when the data is used for archival purpose, the data can be
  located at a network node specialized in storing the neighbors
  collected data. In this way, when a query is issued to the sensor
  network, the query processor will collectively verify every storage
  node for the collected data.
  \item \textbf{Centralized Query Processing}: a leading network sink receives
  the collected data from the peer nodes. In this case, the query processor
  queries the solo centralized node for the global values of the
  participating sensor devices.
\end{itemize}

There are pros and cons related to both of the approaches of query processing,
which are also related to the infrastructure, sensor network nature and
requirements. The primary purpose of the in-network query processing is to
minimize the number of data transmitted from the sensor devices to the final
data sink, as sensor devices are generally limited by constraints such as
energy utilization \cite{sn-storage03}. Along with this strategy, the literature
describes different proposals that aim at upgrading the Structured Query
Language (SQL) adding location-sensitive clauses to the language \cite{sn-db-newop}.
However, this crucial problem related to this method is that it may potentially
flood all nodes with query messages \cite{sn-storage04}.

In order to mitigate the problems related flooded nodes in in-network query
processing, the use of a Data-Centric storage strategy has been shown to be a
proven alternative for the purpose of energy consumption
\cite{sn-storage03, sn-storage01}.Its primary goal is to organize the
collected data based on a particular property, say location or event, in order
to be sent to individual storage nodes that relate to the observation. As the
main outcome of this strategy is to create network partitioning strategy, the
query processor would hit specific sensor nodes with the needed data and,
therefore, would decrease the load and maximize the capacity on the actual
sensor devices.

A centralized network query processing processes data in a network node
specified as the network sink. The network sink usually is a network node with
a database server with the collection of data delivered by all the sensor
devices from the network. As a consequence of this many-to-one communication
strategy \cite{sn-storage02}, the query process does not affect any of the
sensor nodes in the network and, thus, maximizes the sensor device
utilization. Moreover, the use of the data for historical purposes is easier
to manage and provide access to the primary users of the data, following the
process of data processing, compression, aggregation, etc
\cite{sn-db-modeling02}.

Although this approach seemed to be appropriate to the point of view of the
sensor nodes and used as the tradeoff between achieving capacity and energy
consumption, the centralized data collector may suffer from the unavoidable
development of a point of traffic concentration \cite{sn-storage02}, also
referred to as the phenomenon of congestion, or funneling effect
\cite{sn-storage04}. For this reason, simple distributed system techniques
such as the use of master-slave or data replication can be used to reduce the
overhead.

@@

Different approaches are used to represent the collected data from sensors. The
use of basic data descriptors has been used in research groups which expertise
is not database systems, while others use such technology and the way to
represent data using more pragmatic approaches. \cite{sn-data-model-survey}
have shown different data models and the advantages and disadvantages. The 
following is a overview of these approaches used by the sensor network
community.

\begin{itemize}
  \item \textbf{Tabular Data Model}: comma-separated values are immensely
  popular due to its simplicity. Although data is not index , it is an easy
  approach for data exchange among researchers \cite{sn-provenance};
  \item \textbf{Relational Data Model}: it is used in the majority of the
  projects. Most of them due to the easy access to database systems, with
  refactored versions that support the implementation of sensor networks and
  its requirements \cite{sn-db-tinydb};
  \item \textbf{Structured Data Model}: similar to the relational model,
  structured data models are becoming popular due to the Internet. The use of
  XML data \cite{xml} is not yet popular, but there are a few projects using
  it \cite{sn-xml-usage01}\cite{sn-xml-usage02}, implementing its query by
  using a middleware \cite{sn-xml-middleware, sn-xml-query-engines}.
\end{itemize}

@@

Simple use of Comma-Separated Values is an example of sensor network data. Data
is structured by the top header, and the values are in each line of the file.
The implementation of the SF-BEAMS \cite{sfbeams2006} uses this approach. In
summary, in a tabular data model

\begin{itemize}
  \item data is represented by columns in a flat file separated by a character
  that separates the values. One example of such data in sensor networks is the
  OPenDAP \cite{opendap} format, used by the SF-BEAMS project
  \cite{sfbeams2006};
  \item data is queried using a text editor file, without any indexing
  associated with it.
\end{itemize}

@@

Many different projects have used the relational model \cite{relational-model}
as the representation of data for sensor networks. In general, the following
properties apply to sensor networks:

\begin{itemize}
  \item data is structured in tables, and sensor properties are represented by
  the columns of these tables. Data instances are represented by tuples or rows
  of these tables. Finally, table relationships are implemented based on the
  relational algebra theory, which is the mechanism used to describe the
  relationship among different data types.
  \item data is queried using the Structured Query Languange (SQL) and uses the
  definitions of the tables and columns to extract data.
\end{itemize}

Before the creation of such a model, a database normalization process is
necessary to describe the data in a convenient way. An example of such use of
the relational model is the TinyDB database systems \cite{sn-db-tinydb}, which
uses the relational model to represent the types of sensors. This is a popular
choice because of the vast availability of research on relational database
models.

Although the implementation of a relational database seems to cover the needs
of data representation, \cite{sn-data-model-survey} described the problems
related to this approach. One such problem is that the relational model is not
suited for modeling time series data and metadata related to identify, time
and location. For problems related to it, came the implementation of different
approaches that add semantics to data types and different query clauses to the
Structured Query Language (SQL) to solve join operations \cite{sn-db-newop}.

@@

The use of XML is used to represent data in sensor networks has become popular
due the popularity of this data format. In general, the use of XML for
persistence can be summarized as follows:

\begin{itemize}
  \item data is structured in the collected data properties can be represented
  through the XML \cite{xml} instances by using tags, having its model validated
  by the use of XML Schema \cite{xml-schema}.
  \item in order to extract data from XML-based databases, the use of XML XPath
  \cite{xml-xpath} is used to query XML documents.
\end{itemize}

Some examples of the application of XML in sensor networks in
\cite{sn-xml-usage02} some have investigated the effective use of XML in
sensor networks to represent the collected data \cite{sn-xml-usage01}. An
example of querying XML data in sensor networks described by
\cite{sn-xml-query-engines}, while others use XML as a means of data exchange
by implementing of middleware to query and extract query data
\cite{sn-xml-middleware}. NetBEAMS is one of such sensor network that contains
an XML-based communication middleware for data exchange among the sensor nodes
\cite{netbeams2009}.

@@

Most of sensor networks implementations use the collected data from the sensor
devices after they perceive the environment. As shown by
\cite{sn-data-model-survey}, different data models contains their pros and
cons to represent data. However, one crucial feature in sensor networks is the
description of it collected data, as \cite{sn-provenance} discussed the
properties and nature of sensor networks generated data. First, the device's
sampling are usually in raw format or without obvious names. In order to
provide search over the collected sensor data, the data needs first to be
described and indexed, and this process can be related to Data Provenance
\cite{db-provenance}. Needless to say, Data Provenance is the
Metadata\footnote{Data about data} about the collected sensor data.

An example of Data Provenance application is the use of aggregation of data
over time to estimate the changing effects of the weather in a given region.
For this reason, the collected data needs descriptions or annotations that can
provide enough information to track changes, that is, the history of how and
when a given data has changed. The following should be the essence of a
provenance-aware collected data according to \cite{sn-provenance}.

\begin{itemize}
  \item \textbf{what was collected?}: it can be merely one or more low-level
  observations,  which is a set of values with or without a description;
  \item \textbf{When was the data collected?}: is the time when the data was
  collected,  as most sensor data must be used depending on its existence;
  \item \textbf{Where was the data collected?}: it is typically the value of 
  the of the location from where the data was collected. However, it is vital to
  note that the use of this metadata depends on the applicability of the data.
\end{itemize}

While Data Provenance provides basic mechanisms to describe sensor networks
data in general, others see the importance of the time space of sensor
networks. \cite{sn-time-series} describes strategies to used to cluster
distributed sensor data as time series of data, as well as to decrease the
size of data streams being transmitted in real-time sensor networks when these
collected data does not need to be transferred \cite{sn-data-reduction}.

Finally, an indispensable application of collected sensor data is the so-called
blackbox of an aircraft, which carries the collection of raw data collected
from a group of sensors placed in different places of the aircraft, as
described by \cite{sn-exemple-blackbox}. The author describes the data is used
in a process of data recovery and analysis in order to identify the probable
reason of an aircraft accident, by using the time, and location of the
aircraft, along with the independent variables from the sensors.

>>>>

In general, this chapter reviews surveys and related papers in networks
networks, providing a basic introduction about their characteristics to new 
audience. Similarly, it reviews the characteristics of related works that
discusses and provides data persistence solutions for sensor networks, 
characterizing the different approaches used to described and model the
gathered data from sensor devices.

\section{Survey in Sensor Networks and Its Infrastructure}

According to \cite{sn-intro02}, sensor networks are specialized types of
network systems comprised of nodes with specific goals such as to produce and
collect data. The former is seen as a sensor device capable of producing data
to the network based on an observation from the environment or simply a raw
data. On the other hand, the latter is described as the final destination of
the raw data in a sensor network and called the network data sink, or simply
data sink, whose responsibility is to receive, categorize and store the
``sensed'' data. In this way, the location of the the data sink is determined by
the purpose of the collected data: data nodes are placed in-network close to
the sensor devices when collected data has a life cycle, being discarded after
being utilized. In contrast, data nodes are placed in in a centralized node
when the collected data is used for historical purposes.

The following sections describe general properties of sensor networks regarding
infrastructure and components as described in the literature.

\subsection{Properties of Sensor Device Nodes in Sensor Networks}

Each sensor node perform different tasks for the network and, therefore,
influences on the method of deployment used. \cite{sn-intro01} identifies that
the deployment of sensor devices can use arbitrary methods such as being
dropped from an airplane, or manually placed at chosen geographic locations.
Moreover, those authors also mentioned that the deployment can be done as a
one-time event or as an iterative process, where the sensor devices are moved
to different places as needed.

Moving to a differnt point, \cite{sn-intro01} reports that sensor nodes can be
able to move around the place where they were deployed or a stationary sensor
at a given location. In the case of a sensor networks comprised with different
moving sensors, a given sensor node can move occasionally or continuously, as
a result of an accidental side-effect or a desired property of the device. As
a matter of fact, these sensors are automotive or they can be attached to
another moving object. For this reason, \cite{sn-intro01} classifies those
types of mobility as active or passive.

Considering the physical design and capabilities of sensor devices, they range
from microscopically small to bigger devices of a few cubic feet
\cite{sn-intro02}. In addition to their limited size, their computational
resources and power availability are also scarse, when they are located in
areas of limited access to energy. In this way, \cite{sn-intro01} also
distinguishes how sensor devices are usually designed to either store energy on
batteries or use energy scavenged from solar cells. Even though a sensor
network is designed for a purpose, \cite{sn-intro01} also explains that it can
be comprised of different sensor devices playing different roles on what they
observe in the environment, characterizing the heterogeneity the sensor
network by the hardware capabilities and what types of data each of them
produce.

Finally, \cite{sn-intro01} mentions that sensor device nodes can communicate
with each other using different media such as radio, diffuse light or
inductive sound, influencing on the type of protocols they exchange data. They
can have knowledge of their physical location if they are integrated with a
GPS\footnote{Global Positioning System} device, which gives the latitude or
longitude coordinate values. Therefore, their cost of these devices may range
from hundreds to thousands of dollars, depends on a given device's functionality
and capabilities of its design.

\subsection{Properties of the Infrastructure of Sensor Networks}
\label{sec:sn-infrastructure}

Last section discussed the different properties of the sensor devices when
placed in a sensor network. Depending on how they communicate with each other,
\cite{sn-intro02} discusses that sensor devices can be in an organized in a
structured or ad hoc way. For this reason, the infrastructure, of sensor
networks differ in topology such as single-hop, star, tree or  graph. Each of
these topologies may influence on how the produced data from sensor nodes is
collected and routed to their related data sink. While there are only two
levels of node connectivity in sensor networks deployed as single-hop and start
topologies, sensor networks whose topology is either tree or graph use
multi-hop communication among each of their nodes for data communication from
the devices to the data sink.

The property of coverage on a sensor network is defined by
\cite{sn-intro01} as the physical space ``visible" to the sensors devices, which
is also directly related to the purpose of the network, its size in terms of
the number of nodes deployed and the sensor network lifetime. Therefore,
sensors are deployed in sparse or dense manners, according to the
specification of what needs to be covered by the sensor devices. Furthermore,
considering that some types of sensor devices are very likely to regular
hardware failures, \cite{sn-intro2} described sensor network examples with 
redundant nodes that provide better data coverage and levels of coverage.
Finally, the coverage of sensor devices also influence on the connectivity
property of the sensor nodes. In order communicate with redundant nodes, a
physical connection must be available at all times. On the other hand, lower
level of coverage is is related sporadic communication among the sensor nodes.

The sensor network size is related to the degree of coverage mentioned earlier,
and it is measured by the number of devices deployed. \cite{sn-intro01}
reports examples of sensor networks composed by different ranges of deployed
sensor devices such as dozens, hundreds or even thousands of nodes. However,
the life time of a sensor network determines how long these devices are going
to be used, as well as how long the sensor network may exist. For example, an
environmental sensor network \cite{sn-ex01} with the purpose of collecting data
for historic purposes may be have their sensor devices continuously performing
their data collection tasks. On the other hand, however, the lifetime of a
sensor network designed to achieve a goal may not exist after the sensor
devices finish their tasks. Finally, \cite{sn-intro02} argues that the
performance related to the orchestration of sensor devices in sensor networks
is related to the constraints defined for quality-of-service (QoS) such as
real-time data delivery within a thredshold or for robustness such as
maintaining data alive even though a communication link is not available.

All things considered, sensor nodes play an important role during the execution
of a sensor network, whose infrastructure is characterized by how and where the
sensor nodes are deployed and communicate with each other. In this way, next
section covers the approaches to deal with the product of this type of system
network.

@@

As traced in the last few sections, sensor networks are comprised of
different types of sensors nodes. However, an important sensor node type is the
data sink because it is the place where the raw data produced by sensor
nodes resides. Data collection and access in sensor networks is directly
related to the characteristics of the sensor network, its routing mechanisms
and where the data sink is placed within the network. In a nutshell,
\cite{sn-storage03} describes collected data as can be described as Low-level
property or an observation such as the level of the ocean, etc.

The type of query mechanism used to directly or indirectly interrogate sensor
devices in a sensor network is directly related to the aforementioned locations
of the collected data. The in-network query processing is used when the
location of the collected data is a local storage \cite{sn-storage01,
sn-storage03} located in the network, strategy typically used to maximize the
sensor's resources, while minimizing energy consumption \cite{sn-storage04}. On
the contrary, the a typical centralized query processing is used to search for
values over the collected data when its location is a centralized data
node \cite{sn-storage02}.

When it comes to the data representation, the relational data model
\cite{relational-model} is one of the most traditional approaches used to
describe the collected data of sensor networks, as it is reported in database
systems such as the TinyDB \cite{sn-db-tinydb}. For instance, \cite{sn-db-newop}
proposes a modified version of the SQL language of TinyDB proposed to solve
SQL Join problems, so that the query processing takes into account the location
of the distributed data. Conversely, \cite{sn-provenance} proposes the use of
Data Provenance as the primary guidelines to describe collected data from
sensor devices, giving the fact that these devices only produce observations as
stream of raw data, and therefore, described to address the problems related to
the origin of the data.

The following sections describe general properties of sensor networks regarding
data persistence, as reviewed in the literature.

\subsection{Data Use and Where Data is Stored}
\label{sec:sn-data-purpose}
\label{sec:sn-storage-locations}

At first glance, the raw data serves as the main outcome of the sensor network.
\cite{sn-provenance} differentiates the use of the collected data as follows as 
\textbf{real-time data stream}, when they are generated data from a sensor
device whenever the data is requested. In this way, the data is still in its raw
format specified by the sensor manufacturer; On the other hand, the authors
describes the data as \textbf{archivable}, since they a have historical
significance and will be processed to be reused by other systems in a latter
time. While the real-time data streams are used and potentially discarded, the
archivable data is stored into secondary memory such as a hard-disk. Under
these circumstances, the collected data needs first to face its journey
starting from its production, transmission from the sensor device, to finally
reach a specified data sink in the network \cite{sn-storage01, sn-storage02,
sn-storage03}.

After producing the ``sensed'' values for the observed parameters from the
environment, the sensor node is responsible to send the raw data to a specific
data node defined by the network infrastructure, as delineated in section
\ref{sec:sn-infrastructure}. In such a way, \cite{sn-storage03}
describes that the location for the collected sensor data provides two
different types of storage devices: the \textbf{external storage} and the
\textbf{local storage}. In the former type, the sensor device sends the
produced data to another sensor node that contains an external storage, whereas
the data only exists in memory or between its reading to its delivery as
real-time data stream in the the latter type. On the whole, the sensor network
infrastructure dictates where the movement performed by the sensor data, as
\cite{sn-storage02} sees a wireless sensor networks as a many-to-one data
collection pattern, when its organization is based on single-hop or star. On
the other hand, data may also flow from one intermediate sensor node to another
until it reaches its final data sink when the data organization is structured
as a graph, tree \cite{sn-storage01, sn-storage03}.

Last, but not least, the use of different external storage devices can be used
to partition the collected data. This practice is called \textbf{Data-Centric
Storage}, and aims at clustering the collected data from different sensor
devices based on a given a property of the data such as location or time of
creation.

\subsection{Query Processing in Sensor Networks}
\label{sec:query-process}

The query processing in sensor networks is a direct or indirect method of
accessing the collected data stored at data nodes, whose location is determined
by different facts described in the previous section. \cite{sn-storage03}
describes an example of an \textbf{in-network query processing} when sensor
nodes, acting as data nodes, provides real-time data stream as a response
to queries requesting the device's collected data. Likewise, if the collected
data is in use for a specific period of time, the data is cached in an external
storage device in a close-by data node. In this way, when a query is issued to
the sensor network with these characteristics, the query processor considers 
all data nodes return its final response to the query. In comparison, a sensor
network whose topology is in form of star, the data sink is usually located in
the center of structure, and thus, receives all the collected data from other
sensor devices in the network. For this reason, it can be seen as a 
\textbf{centralized query processing}. Factoring in that a centralized data
sink receives all the collected data from the peer nodes, this query processing
strategy is the solo responsible for responding to indirect queries about the
produced data from the network.

Based on the literature review, there are advantages and disadvantages related
each of the query processing approaches described with regards to the
infrastructure of the sensor network. \cite{sn-storage03} argues that the 
primary purpose of the in-network query processing is to minimize the number
of data transmitted from sensor devices to a data sink, as sensor devices are
generally limited by energy use. Considering that sensor devices are also used
as data nodes, \cite{sn-storage04} delineate that this approach can potentially
flood all sensor nodes in the network, and therefore, decrease its performance.
For this reason, \cite{sn-storage01} reports the use of the data-centric
storage by organizing the collected data based on its particular properties
values. In this way, the query processor hits specific data nodes that contains
the types of data in the range of the requested data and, therefore, the author
concludes that this approach decreases the network congestion and maximizes the
capacity on the sensor devices and the performance of the sensor network.

Unlike the in-network, the centralized query processing is related to a regular
database system, located in one of the network nodes. For this reason, this
indirect type of query processing is used when the data purpose of the use of
data is for data archival. However, \cite{sn-storage02} showes that as a
consequence of the many-to-one communication between the centralized data sink
and the other data nodes, this query process does not affect any of the sensor
nodes in the network and, thus, maximizes the sensor device utilization. 
Moreover, this strategy better exposes the collected data to the primary users
raw data, which follows the process of data processing, compression,
aggregation, and so on \cite{sn-db-modeling02}.

Although the approach of centralized query processing maximizes the sensor
nodes capacity and minimize energy consumption, \cite{sn-storage02} explains
that this trade off imposes the creation of the unavoidable development of a
point of traffic concentration or the phenomenon of network congestion. For
this reason, \cite{sn-storage04} describes that different techniques used in 
distributed system such as the use of master-slave data replication or 
database partitioning \cite{db-partitioning-relational} can be used to reduce
or minigate the effects of this funneling effect.

@@

Different approaches are used to represent the collected data from sensors. The
use of basic data descriptors has been used in research groups which expertise
is not database systems, while others use such technology in pragmatic
approaches. \cite{sn-programming-language} reports different applications using
different types of languages to acces the data modeled in different data
models. \cite{sn-provenance} describes the use of a \textbf{Tabular Data Model}
\cite{tabular-model} using comma-separated values to model and organized the
values related to the properties of observed parameters from a given sensor
device. On the other hand, \cite{sn-db-tinydb} describes the use of the
\textbf{Relational Data Model} \cite{relational-model} is used in the majority
of the projects that involves data persistence due to its popularity in other
different areas of Software Engineering and Distributed Systems. Finally,
\cite{sn-xml-usage01, sn-xml-usage02} are examples of sensor networks
application that makes use of XML \cite{xml} to represent the collected data
on a \textbf{Structured Data Model}.

The following sections describe the application of each of aforenamed data
models in the context of sensor networks.

@@

The most basic computer application to process data is through the use of
spreadsheets, where data is organized into cells \cite{tabular-model}. When
exported to the so-called comma-separated values (CSV) file format, the
resulting file is comprised of lines of ascii characters, separated by a
delimiter. The first line represents a header, with the identification of
columns of specific data, where the following lines contains the values
related to each top column. \cite{sn-provenance} reports that the tabular
data model is commonly used to provide data persistence for sensor networks.
As a matter of fact, the SF-BEAMS environmental sensor networks
\cite{sfbeams2006} uses this approach to distribute the collected data over
the Internet using the OPenDAP \cite{opendap} data format.

This data model can be used in different context, as \cite{sn-provenance}
describes that the versions of the same document can be easily managed and
shared among research colleages using a version-control system such as
Subversion \cite{subversion}.

@@

\cite{sn-db-newop} shows the use of the Relational Data Model
\cite{relational-model} can be used in sensor networks to describe the
collected data from sensor devices. As a matter of fact, the project described
by the authors is based on the TinyDB database system \cite{sn-db-tinydb}, a
database system specifically built for data nodes of sensor networks. It uses
the same properties and constraints of regular database system, where the
entities of the sensor network are required to be modeled and represented by
tables.

In order to query the data collected from sensor devices, the use of the 
Structured Query Language (SQL) \cite{sql} is the main language used to extract
the data managed by the system. As mentioned earlier, the default versions of
this language have been modified to give support to the constraints of the
location of the database such as the addition of new SQL operators for TinyDB
\cite{sn-db-newop}.

@@

XML is an increasing data format used popularized by the applications developed
for the Internet. \cite{sn-xml-usage01} reports an efficient way to represent
the collected data from sensors using XML, while the authors of
\cite{sn-xml-usage02} describe a sensor network using XML for data model
representation. In general, the collected data is structured into classes of
relating elements and validated by an XML Schema \cite{xml-schema}.

In order to extract data from XML-based databases, the XML XPath language
\cite{xml-xpath} is often used to query XML data. However, with the inception
of hybrid XML-relational database, \cite{sn-xml-usage03} reported the use XML
as the data model and the use of SQL to query the stored data on the IBM's DB2
database system.

Other examples of the use of the structured data model using XML is the through
the use of middlewares, as the authors of \cite{sn-xml-query-engines} reports.
Such approach is used for data exchange by the implementing of a middleware
capable of query and extract the collected data from sensor devices
\cite{sn-xml-middleware}. A practical example of the use of XML as a middleware
to communicate between nodes is the communication mechanism used by NetBEAMS
\cite{netbeams2009}. See section \ref{sec:dsp-serialization} for details.

@@

As described in the previous section, the collected data from sensor networks
can be represented by different data models, requiring the knowledge of the
properties of the data. Not to mention that the low-level or raw data produced
by a given sensor device might not include vital information regarding the
origing and identity of the collected data. For this reason,
\cite{sn-provenance} describes Data Provenance as a technique used to describe
those essential properties of the collected data from sensors. One typical
example of the application of Data Provenance is the so-called blackbox of an
aircraft, which carries the collection of raw data collected from a group of
sensors placed in different locations of an aircraft, as shown by
\cite{sn-exemple-blackbox}. Last, but not least, the author describes the
collected data as used in a process of data recovery and analysis in order to
identify the probable reason of an aircraft accident, by using the time, and
location of the aircraft, along with the independent variables from the
sensors.

In general, sensor devices produce data and either keep them onto its local
storage or transmit them to data nodes as described earlier in this chapter. It
is vital to understand that the majority of these devices outputs the data in
its raw format, containing only a stream of values with any type of
delimitation between each other. For this reason, the processes of parsing and
description of each of the values located in the data streams regularly require
the use of the device's specification documentation. Similarly, in order to
have access to each of the properties of data collected by a given property
related to time or space, \cite{sn-provenance} also mentions that 
Metadata\footnote{Data about data} should be included together with the
collected data.

An example of Data Provenance application is the use of aggregation of data
over time to estimate the changing effects of the weather in a given region.
For this reason, the collected data from sensor devices need meaningful
annotations to better describe the identity of the collected data, as 
\cite{sn-provenance} reports. Interesting questions regarding the
data can be asked such as \textbf{what was collected?}, \textbf{When was the
data collected?} or \textbf{Where was the data collected?}. Similarly, while
Data Provenance provides basic mechanisms to describe sensor networks data in
general, others see the importance of the time space of sensor networks.
\cite{sn-time-series} describes strategies used to cluster distributed sensor
data as time series of data, as well as to decrease the size of data streams
being transmitted in real-time sensor networks when these collected data does
not need to be transferred \cite{sn-data-reduction}.

Finally, data provenance provides practical use of the collected data from
different sensor devices and different sensor networks in the process called
Information Fusison \cite{sn-info-fusion}. An example of the aggregation of
different data from different devices located in different geographic sensor
networks is \cite{sn-geo-metadata}, which is capable of process
spatial-temporal queries of real-time or historical data. The following
sections describe each of the metadata properties as defined by
\cite{sn-provenance}, giving different examples of each of them.
